### 2017-11-27 submission/manuscript.Rmd
fix a final typo
update with Pat's comments 3-6

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces, but not tissue, were  lower in those with carcinoma [Figure 1]. Using linear mixed-effect models to control for study and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). However, in tissue, this effect was not observed in tissue when resampling of the same individual was also controlled for (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). Within fecal samples, a decrease in Shannon diversity and evenness resulted in a significantly increased RR for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2]. Only a decrease in evenness had an increased RR for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).

Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S1 & S2]. For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S3 & S4].

***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. In fecal samples the RR due to CRC associated genera was greater than either the RR assoicated with evenness or Shannon diversity [Figure 2 & 3]. Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual.

@@

***Using the Whole Community Instead of Only CRC-Associated Genera Increases Model AUC:***
For both fecal and tissue (matched and unmatched) samples, the AUC decreases when only OTUs from the CRC-associated genera are used [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5]. The genus models generally performed similar to the OTU-based models with the full genera models performing better than the CRC-associated genera models [Figure S2-S3]. Both genus models were similarily predictive in their ability to detect adenomas or carcinomas, with carcinoma having a higher AUC then adenoma [Figure S4-S5]. Of note, matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] than carcinoma models [Figure S4B & S5A].

***A Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B]. No single study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 6A & B].  In order to achieve adequate power for small effect sizes, studies would need to recruit over 1000 individuals for each arm [Figure 6C].

@@

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. $\beta$-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched) tissue samples. Where applicable for each study, predictions for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models, the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power.

@@

**Figure 6: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with adenoma individuals. B) Power based on effect size for studies with carcinoma individuals. C) The estimated sample number needed for each arm of each study to detect aneffect size of 1-30%. The dotted red lines in A) and B) represent a power of 0.8.

@@

**Figure S4: Random Forest Prediction Success Using Genera for each Stool Study.** A) AUC for prediction in adenoma using all genera or CRC associated genera only. B) AUC for prediction in carcinoma using all genera or CRC-associated genera only. The dotted line represents an AUC of 0.5. The x-axis is the data set in which the model was initially trained on.

**Figure S5: Random Forest Prediction Success of Carcinoma Using Genera for each Tissue Study.** A) AUC for prediction in unmatched tissue for all genera or CRC-associated genera only. B) AUC for prediction in matched tissue using all genera or CRC-associated genera only. The dotted line represents an AUC of 0.5. The x-axis is the data set in which the model was initially trained on.

**Figure S6: Random Forest Prediction Success of Adenoma Using Genera for each Tissue Study.**

>>>>

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
Using the combined data set we first assessed whether there were any broad scale community differences that could be detected as disease severity worsened. Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces, but not tissue, were  lower in those with carcinoma [Figure 1]. Using linear mixed-effect models to control for study and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). However, in tissue, this effect was not observed when resampling of the same individual was also controlled for (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). We next tested whether these detectable differences in the community resulted in a significant increase in RR. Within fecal samples, a decrease in Shannon diversity and evenness resulted in a significantly increased RR for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`) and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)) [Figure 2]. Only a decrease in evenness had an increased RR for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)). Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].

Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S4 & S5]. For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S6 & S7].

***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
Based on the small increase in RR using $\alpha$-diversity metrics, we wanted to know if the presence of specific genera resulted in a higher RR for both stool and tissue. To investigate this we analyzed the classically associated CRC genera, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for and increase in RR. The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. In fecal samples the RR due to CRC associated genera was greater than either the RR assoicated with evenness or Shannon diversity [Figure 2 & 3]. Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual.

@@

***Using the Whole Community Instead of Only CRC-Associated Genera Increases Model AUC:***
We then tested whether the overall bacterial community was at all important to classifying disease or if the CRC-associated genera were sufficient alone. To test this we used two approaches. The first used genus level data and tested whether there were any differences in AUC when training on one study and testing on all the others when using either all genera present or only the CRC-associated genera. The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model across studies using either all OTUs or only OTUs that taxonomically classified to CRC-associated genera.

The genus based models showed an AUC decrease in model classification on the training set for both stool and tissue studies [Figure S2-S3]. With respect to the test sets, comprised of genera data from other studies, both the all genera model and CRC-associated models had a similar ability to detect adenomas or carcinomas [Figure S4-S6]. The AUC for classification of adenomas was lower than carcionmas for both tissue and stool [Figure S4-S6].

The OTU based models for both fecal and tissue (matched and unmatched) samples, showed an AUC decrease when only OTUs from the CRC-associated genera are used versus full community of OTUs [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5] with the largest difference in median AUC for stool being in carcinoma classification [Figure 4B] and for tissue being in adenoma [Figure 5A].

***A Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
Based on the previous observations we then assessed whether the studies that we included are realistically powered to identify small, medium, and large scale differences between case and control. When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B]. No single study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 6A & B].  In order to achieve adequate power for small effect sizes, studies would need to recruit over 1000 individuals for each arm [Figure 6C].

@@

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. $\beta$-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched) tissue samples. Where applicable for each study, predictions for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models, the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power. For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study. Similarily, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.

***Obtaining CRC-Associated Genera:***
For the CRC-associated genera analysis of the RR, the total average counts were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for 100 different subsamplings. The OTU based Random Forest Models that used CRC-associated genera used a similar approach except that the OTUs were not aggregated together by genus by kept as separate OTUs. So, OTU Random Forest models using the full community included all OTUs while those using CRC-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples that had control and lesion samples that did not belong to the same indivdiual were classified as unmatched while samples that belonged to the same individual were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu. Studies that had unmatched data were Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched due to one of the corresponding matched samples not making it through sequence processing. For the linear mixed-effect models samples from the same individual were taken into account. For all other analysis matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.

@@

**Figure 6: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with adenoma individuals. B) Power based on effect size for studies with carcinoma individuals. C) The estimated sample number needed for each arm of each study to detect an effect size of 1-30%. The dotted red lines in A) and B) represent a power of 0.8.

@@

**Figure S4: Random Forest Prediction Success Using Genera for each Stool Study.** A) AUC for prediction in adenoma using all genera or CRC associated genera only. B) AUC for prediction in carcinoma using all genera or CRC-associated genera only. The dotted line represents an AUC of 0.5. The x-axis is the data set in which the model was initially trained on. The red lines represent the median AUC using that specific study as the training set.

**Figure S5: Random Forest Prediction Success of Carcinoma Using Genera for each Tissue Study.** A) AUC for prediction in unmatched tissue for all genera or CRC-associated genera only. B) AUC for prediction in matched tissue using all genera or CRC-associated genera only. The dotted line represents an AUC of 0.5. The x-axis is the data set in which the model was initially trained on. The red lines represent the median AUC using that specific study as the training set.

**Figure S6: Random Forest Prediction Success of Adenoma Using Genera for each Tissue Study.** The red lines represent the median AUC using that specific study as the training set.

### 2017-11-27 submission/manuscript.tex
fix a final typo
update with Pat's comments 3-6

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} Using power transformed and Z-score normalized
\(\alpha\)-diversity metrics, both evenness and Shannon diversity in
feces, but not tissue, were lower in those with carcinoma {[}Figure
1{]}. Using linear mixed-effect models to control for study and variable
region, there was a significant decrease from control to adenoma to
carcinoma for both evenness (P-value = 0.025) and Shannon diversity
(P-value = 0.043). However, in tissue, this effect was not observed in
tissue when resampling of the same individual was also controlled for
(P-value \textgreater{} 0.05). Within fecal samples, a decrease in
Shannon diversity and evenness resulted in a significantly increased RR
for carcinoma (P-value = 0.01 and 0.0011, respectively) {[}Figure 2{]}.
Although these values were significant, the effect size was relatively
small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
{[}Figure 2{]}. Only a decrease in evenness had an increased RR for
adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect size was
even smaller than what was observed for carcinoma (RR = 1.16).

Using the Bray-Curtis distance metric, there was a significant
difference across studies in the bacterial community of fecal samples
between carcinoma and controls, but not adenoma and controls {[}Table S1
\& S2{]}. For studies with unmatched tissue samples a similar trend was
observed {[}Table S3 \& S4{]} while studies with tissue samples from the
same individual (matched) had no differences {[}Table S3 \& S4{]}.

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} The majority of CRC-associated genera for both feces and
tissue had a significantly increased RR for carcinoma but not for
adenoma {[}Figure 3{]}. In fecal samples the RR due to CRC associated
genera was greater than either the RR assoicated with evenness or
Shannon diversity {[}Figure 2 \& 3{]}. Additionally, the RR of carcinoma
continuously increased as individuals tested positive for more CRC
associated genera {[}Figure 3B \& 3D{]}. The RR effect size was greater
for stool (RR range = 1.61 - 2.74) than for tissue (RR range = 1.21 -
1.81). This decrease may be explained by the fact that the tissue
analysis included matched samples from the same individual.

@@

\textbf{\emph{Using the Whole Community Instead of Only CRC-Associated
Genera Increases Model AUC:}} For both fecal and tissue (matched and
unmatched) samples, the AUC decreases when only OTUs from the
CRC-associated genera are used {[}Figure 4 \& 5{]}. This decrease is
observed in both adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The
genus models generally performed similar to the OTU-based models with
the full genera models performing better than the CRC-associated genera
models {[}Figure S2-S3{]}. Both genus models were similarily predictive
in their ability to detect adenomas or carcinomas, with carcinoma having
a higher AUC then adenoma {[}Figure S4-S5{]}. Of note, matched tissue
samples for those with carcinoma had an AUC that was more similar to the
adenoma models {[}Figure S4A, S5B, \& S6{]} than carcinoma models
{[}Figure S4B \& S5A{]}.

\textbf{\emph{A Majority of Studies are Underpowered for Detecting Small
Effect Size Differences:}} When assessing the power of each study at
different effect sizes the majority of studies for both adenoma and
carcinoma have an 80\% power to detect a 30\% difference {[}Figure 6A \&
B{]}. No single study that was analyzed had the standard 80\% power to
detect an effect size difference that was equal to or below 10\%
{[}Figure 6A \& B{]}. In order to achieve adequate power for small
effect sizes, studies would need to recruit over 1000 individuals for
each arm {[}Figure 6C{]}.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
\(\beta\)-diversity was then assessed for each inidividual study. Next,
four specific CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
were assessed for differences in relative risk. We then built Random
Forest models based on all genera or the select CRC-associated genera.
The models were trained on one study then tested on the remaining
studies for every study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched) tissue samples. Where applicable for each study,
predictions for adenoma and carcinoma were tested. This same approach
was then applied at the OTU level with the exception that instead of
testing on the other studies, a 10-fold cross validation was utilized
and 100 different models were created based on random 80/20 splitting of
the data to generate a range of expected AUCs. For OTU based models, the
CRC associated genera included all OTUs that had a taxonomic
classification to \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}. The power of each
study was assessed for an effect size ranging from 1\% to 30\%. An
estimated sample n for these effect sizes was also generated based on
80\% power.

@@

\textbf{Figure 6: Power and Effect Size Analysis of Studies Included.}
A) Power based on effect size for studies with adenoma individuals. B)
Power based on effect size for studies with carcinoma individuals. C)
The estimated sample number needed for each arm of each study to detect
aneffect size of 1-30\%. The dotted red lines in A) and B) represent a
power of 0.8.

@@

\textbf{Figure S4: Random Forest Prediction Success Using Genera for
each Stool Study.} A) AUC for prediction in adenoma using all genera or
CRC associated genera only. B) AUC for prediction in carcinoma using all
genera or CRC-associated genera only. The dotted line represents an AUC
of 0.5. The x-axis is the data set in which the model was initially
trained on.

\textbf{Figure S5: Random Forest Prediction Success of Carcinoma Using
Genera for each Tissue Study.} A) AUC for prediction in unmatched tissue
for all genera or CRC-associated genera only. B) AUC for prediction in
matched tissue using all genera or CRC-associated genera only. The
dotted line represents an AUC of 0.5. The x-axis is the data set in
which the model was initially trained on.

\textbf{Figure S6: Random Forest Prediction Success of Adenoma Using
Genera for each Tissue Study.}

>>>>

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} Using the combined data set we first assessed whether
there were any broad scale community differences that could be detected
as disease severity worsened. Using power transformed and Z-score
normalized \(\alpha\)-diversity metrics, both evenness and Shannon
diversity in feces, but not tissue, were lower in those with carcinoma
{[}Figure 1{]}. Using linear mixed-effect models to control for study
and variable region, there was a significant decrease from control to
adenoma to carcinoma for both evenness (P-value = 0.025) and Shannon
diversity (P-value = 0.043). However, in tissue, this effect was not
observed when resampling of the same individual was also controlled for
(P-value \textgreater{} 0.05). We next tested whether these detectable
differences in the community resulted in a significant increase in RR.
Within fecal samples, a decrease in Shannon diversity and evenness
resulted in a significantly increased RR for carcinoma (P-value = 0.0047
and 0.00034, respectively) {[}Figure 2{]}. Although these values were
significant, the effect size was relatively small for both metrics
(Shannon RR = 1.33 (1.09 - 1.62) and evenness RR = 1.36 (1.15 - 1.61))
{[}Figure 2{]}. Only a decrease in evenness had an increased RR for
adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect size was
even smaller than what was observed for carcinoma (RR = 1.16 (1.01 -
1.34)). Interestingly, for both adenoma and carcinoma there was no
increase in RR within tissue samples for any alpha diversity metric
investigated {[}Table S1-S3{]}.

Using the Bray-Curtis distance metric, there was a significant
difference across studies in the bacterial community of fecal samples
between carcinoma and controls, but not adenoma and controls {[}Table S4
\& S5{]}. For studies with unmatched tissue samples a similar trend was
observed {[}Table S3 \& S4{]} while studies with tissue samples from the
same individual (matched) had no differences {[}Table S6 \& S7{]}.

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
metrics, we wanted to know if the presence of specific genera resulted
in a higher RR for both stool and tissue. To investigate this we
analyzed the classically associated CRC genera, \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
for and increase in RR. The majority of CRC-associated genera for both
feces and tissue had a significantly increased RR for carcinoma but not
for adenoma {[}Figure 3{]}. In fecal samples the RR due to CRC
associated genera was greater than either the RR assoicated with
evenness or Shannon diversity {[}Figure 2 \& 3{]}. Additionally, the RR
of carcinoma continuously increased as individuals tested positive for
more CRC associated genera {[}Figure 3B \& 3D{]}. The RR effect size was
greater for stool (RR range = 1.62 - 2.37) than for tissue (RR range =
1.21 - 1.81). This decrease may be explained by the fact that the tissue
analysis included matched samples from the same individual.

@@

\textbf{\emph{Using the Whole Community Instead of Only CRC-Associated
Genera Increases Model AUC:}} We then tested whether the overall
bacterial community was at all important to classifying disease or if
the CRC-associated genera were sufficient alone. To test this we used
two approaches. The first used genus level data and tested whether there
were any differences in AUC when training on one study and testing on
all the others when using either all genera present or only the
CRC-associated genera. The second approach used OTU level data and
tested whether there was a generalized decrease in the 10-fold cross
validation (CV) model across studies using either all OTUs or only OTUs
that taxonomically classified to CRC-associated genera.

The genus based models showed an AUC decrease in model classification on
the training set for both stool and tissue studies {[}Figure S2-S3{]}.
With respect to the test sets, comprised of genera data from other
studies, both the all genera model and CRC-associated models had a
similar ability to detect adenomas or carcinomas {[}Figure S4-S6{]}. The
AUC for classification of adenomas was lower than carcionmas for both
tissue and stool {[}Figure S4-S6{]}.

The OTU based models for both fecal and tissue (matched and unmatched)
samples, showed an AUC decrease when only OTUs from the CRC-associated
genera are used versus full community of OTUs {[}Figure 4 \& 5{]}. This
decrease is observed in both adenoma and carcinoma groups {[}Figure 4 \&
5{]} with the largest difference in median AUC for stool being in
carcinoma classification {[}Figure 4B{]} and for tissue being in adenoma
{[}Figure 5A{]}.

\textbf{\emph{A Majority of Studies are Underpowered for Detecting Small
Effect Size Differences:}} Based on the previous observations we then
assessed whether the studies that we included are realistically powered
to identify small, medium, and large scale differences between case and
control. When assessing the power of each study at different effect
sizes the majority of studies for both adenoma and carcinoma have an
80\% power to detect a 30\% difference {[}Figure 6A \& B{]}. No single
study that was analyzed had the standard 80\% power to detect an effect
size difference that was equal to or below 10\% {[}Figure 6A \& B{]}. In
order to achieve adequate power for small effect sizes, studies would
need to recruit over 1000 individuals for each arm {[}Figure 6C{]}.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
\(\beta\)-diversity was then assessed for each inidividual study. Next,
four specific CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
were assessed for differences in relative risk. We then built Random
Forest models based on all genera or the select CRC-associated genera.
The models were trained on one study then tested on the remaining
studies for every study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched) tissue samples. Where applicable for each study,
predictions for adenoma and carcinoma were tested. This same approach
was then applied at the OTU level with the exception that instead of
testing on the other studies, a 10-fold cross validation was utilized
and 100 different models were created based on random 80/20 splitting of
the data to generate a range of expected AUCs. For OTU based models, the
CRC associated genera included all OTUs that had a taxonomic
classification to \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}. The power of each
study was assessed for an effect size ranging from 1\% to 30\%. An
estimated sample n for these effect sizes was also generated based on
80\% power. For comparisons in which normal versus adenoma were made the
carcinoma samples were excluded from each respective study. Similarily,
for comparisons in which normal versus carcinoma were made the adenoma
samples were excluded from each respective study.

\textbf{\emph{Obtaining CRC-Associated Genera:}} For the CRC-associated
genera analysis of the RR, the total average counts were collected for
each respective OTU that had a genus level taxonomic classification to
\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
\emph{Porphyromonas} for 100 different subsamplings. The OTU based
Random Forest Models that used CRC-associated genera used a similar
approach except that the OTUs were not aggregated together by genus by
kept as separate OTUs. So, OTU Random Forest models using the full
community included all OTUs while those using CRC-associated genera
included only those OTUs that had a genus level taxonomic classification
to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
and \emph{Porphyromonas}.

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples that had control and lesion samples that did not belong
to the same indivdiual were classified as unmatched while samples that
belonged to the same individual were classified as matched. Studies with
matched data included Burns, Dejea, Geng, and Lu. Studies that had
unmatched data were Burns, Flemer, Chen, and Sanapareddy. For some
studies samples became unmatched due to one of the corresponding matched
samples not making it through sequence processing. For the linear
mixed-effect models samples from the same individual were taken into
account. For all other analysis matched and unmatched samples were
analyzed separately using the statistical approaches mentioned in the
Statistical Analysis section.

@@

\textbf{Figure 6: Power and Effect Size Analysis of Studies Included.}
A) Power based on effect size for studies with adenoma individuals. B)
Power based on effect size for studies with carcinoma individuals. C)
The estimated sample number needed for each arm of each study to detect
an effect size of 1-30\%. The dotted red lines in A) and B) represent a
power of 0.8.

@@

\textbf{Figure S4: Random Forest Prediction Success Using Genera for
each Stool Study.} A) AUC for prediction in adenoma using all genera or
CRC associated genera only. B) AUC for prediction in carcinoma using all
genera or CRC-associated genera only. The dotted line represents an AUC
of 0.5. The x-axis is the data set in which the model was initially
trained on. The red lines represent the median AUC using that specific
study as the training set.

\textbf{Figure S5: Random Forest Prediction Success of Carcinoma Using
Genera for each Tissue Study.} A) AUC for prediction in unmatched tissue
for all genera or CRC-associated genera only. B) AUC for prediction in
matched tissue using all genera or CRC-associated genera only. The
dotted line represents an AUC of 0.5. The x-axis is the data set in
which the model was initially trained on. The red lines represent the
median AUC using that specific study as the training set.

\textbf{Figure S6: Random Forest Prediction Success of Adenoma Using
Genera for each Tissue Study.} The red lines represent the median AUC
using that specific study as the training set.
